---
title: "Data 624 - HW1 (Fall 2024)"
author: 'Khyati Naik'
output:
  pdf_document:
    extra_dependencies: ["geometry", "multicol", "multirow", "xcolor"]
  html_document:  
    theme: cerulean
    highlight: pygments
    css: Lab3.css
    toc: true
    toc_float: true
---

```{r,warning=FALSE}
library(fpp3)
```

## 2.1 Explore the following four time series: Bricks from aus_production, Lynx from pelt, Close from gafa_stock, Demand from vic_elec.

### 2.1.1 Use ? (or help()) to find out about the data in each series.

```{r,warning=FALSE}
data("aus_production")
?aus_production

data("pelt")
?pelt

data("gafa_stock")
?gafa_stock

data("vic_elec")
?vic_elec
```

### 2.1.2 What is the time interval of each series?
aus_production is quarterly data from 1956 to 2010.   
pelt is yearly data from 1845 to 1935.    
gafa_stock business day data when the Market is open from 2014 to 2018.    
vic_elec is every 30 minutes data from 2012 to 2014.   

### 2.1.3 Use autoplot() to produce a time plot of each series.

```{r,warning=FALSE}
aus_production %>%   autoplot(Bricks)
pelt %>%   autoplot(Lynx)
gafa_stock %>%   autoplot(Close)
```


### 2.1.4 For the last plot, modify the axis labels and title.

```{r,warning=FALSE}
vic_elec %>%   autoplot(Demand) +
  labs(x = "Date", y = "Demand") +
  ggtitle("Electricity Demand Over Time")
```


## 2.2 Use filter() to find what days corresponded to the peak closing price for each of the four stocks in gafa_stock.

```{r,warning=FALSE}
gafa_stock %>% 
             group_by(Symbol) %>%
             filter(Close == max(Close))
```


## 2.3 Download the file tute1.csv from the book website, open it in Excel (or some other spreadsheet application), and review its contents. You should find four columns of information. Columns B through D each contain a quarterly series, labelled Sales, AdBudget and GDP. Sales contains the quarterly sales for a small company over the period 1981-2005. AdBudget is the advertising budget and GDP is the gross domestic product. All series have been adjusted for inflation.

### 2.3.a You can read the data into R with the following script:

```{r,warning=FALSE}
tute1 <- readr::read_csv("https://raw.githubusercontent.com/Naik-Khyati/data_624/main/hw1/tute1.csv")
View(tute1)
```

### 2.3.b Convert the data to time series

```{r,warning=FALSE}
mytimeseries <- tute1 %>%
  mutate(Quarter = yearquarter(Quarter)) %>%
  as_tsibble(index = Quarter)
```

### 2.3.c Construct time series plots of each of the three series

```{r,warning=FALSE}
mytimeseries %>%
  pivot_longer(-Quarter) %>%
  ggplot(aes(x = Quarter, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y")
```

### 2.3.c.1 Check what happens when you don’t include facet_grid().

```{r,warning=FALSE}
mytimeseries %>%
  pivot_longer(-Quarter) %>%
  ggplot(aes(x = Quarter, y = value, colour = name)) +
  geom_line()
```

When facet_grid() is not included in the ggplot code, the result is a single plot where all the time series data (for different variables represented by name) are plotted together in one chart. 

## 2.4. The USgas package contains data on the demand for natural gas in the US.

### 2.4.a Install the USgas package.

```{r,warning=FALSE}
library(USgas)
```

### 2.4.b Create a tsibble from us_total with year as the index and state as the key.

```{r,warning=FALSE}
ts <- us_total

ts <- ts %>%
  as_tsibble(index = year, key = state)

head(ts)
```

### 2.4.c Plot the annual natural gas consumption by state for the New England area (comprising the states of Maine, Vermont, New Hampshire, Massachusetts, Connecticut and Rhode Island).

```{r,warning=FALSE}
ne_ts <- ts %>%
filter(state %in% c('Maine', 'Vermont', 'New Hampshire', 'Massachusetts', 'Connecticut', 'Rhode Island')) %>%  mutate(y = y/1e3)

head(ne_ts)
```

```{r,warning=FALSE}
autoplot(ne_ts, y) +
  labs(title = "Annual natural gas consumption by state",
       subtitle = "New England",
       y = "Consumption")
```



## 2.5 Tourism Data Analysis

### 2.5.a Download tourism.xlsx from the book website and read it into R using readxl::read_excel().

```{r,warning=FALSE}
library(readxl)
library(httr)

# URL of the raw Excel file
url <- "https://raw.githubusercontent.com/Naik-Khyati/data_624/main/hw1/tourism.xlsx"

# Temporary file to store the downloaded Excel
temp_file <- tempfile(fileext = ".xlsx")

# Download the file
GET(url, write_disk(temp_file, overwrite = TRUE))

# Read the Excel file from the temporary location
tourism_data <- read_excel(temp_file, sheet = "Sheet1")

# View the first few rows of the dataset
head(tourism_data)
```

### 2.5.b Create a tsibble which is identical to the tourism tsibble from the tsibble package.

```{r,warning=FALSE}
# Convert tourism_data to tsibble
tourism_ts <- tourism_data %>%
  mutate(Quarter = yearquarter(Quarter)) %>%
  as_tsibble(index = Quarter, key = c(Region, State, Purpose))

head(tourism_ts)
```

### 2.5.c Find what combination of Region and Purpose had the maximum number of overnight trips on average.

```{r,warning=FALSE}
reg_pur_max_on_trips <- tourism_data %>%
  group_by(Region, Purpose) %>%
  summarise(Trip_Avg = mean(Trips)) %>%
  filter(Trip_Avg == max(Trip_Avg)) %>%
  arrange(desc(Trip_Avg))

head(reg_pur_max_on_trips)
```

The combination of the Sydney region and the purpose of Visiting has the highest average number of overnight trips per quarter, with 747 trips.


### 2.5.d Create a new tsibble which combines the Purposes and Regions, and just has total trips by State.

```{r,warning=FALSE}
reg_pur_tot_trips_state <- tourism_data %>%
  mutate(Quarter = as.Date(Quarter)) %>%
           group_by(State) %>%
           summarise(total_trips = sum(Trips)) %>%
           arrange(desc(total_trips))

head(reg_pur_tot_trips_state)
```

New South Wales, Victoria, and Queensland have a significant lead in total trips.

## 2.8. Use the following graphics functions: autoplot(), gg_season(), gg_subseries(), gg_lag(), ACF() and explore features from the following time series: “Total Private” Employed from us_employment, Bricks from aus_production, Hare from pelt, “H02” Cost from PBS, and Barrels from us_gasoline.

Can you spot any seasonality, cyclicity and trend?
What do you learn about the series?
What can you say about the seasonal patterns?
Can you identify any unusual years?

```{r,warning=FALSE}
data("PBS")
data("us_employment")
data("us_gasoline")
```

```{r,warning=FALSE}
us_employment %>% 
  filter(Title == "Total Private") %>% 
  autoplot(Employed) + 
  ggtitle("Autoplot")

us_employment %>%filter(Title == "Total Private") %>% gg_season(Employed) +
  ggtitle("Seasonal Decomposition")

us_employment %>% 
  filter(Title == "Total Private") %>% 
  gg_subseries(Employed) +
  ggtitle("Subseries Plot")

us_employment %>% 
  filter(Title == "Total Private") %>% 
  gg_lag(Employed) +
  ggtitle("Lag Plot")

us_employment %>%
  filter(Title == "Total Private") %>%
  ACF(Employed) %>%
  autoplot() +
  ggtitle("Autocorrelation Function")
```

The US Employment dataset shows a general upward trend in Total Private employment over the years, with a notable dip around 2008 that aligns with the housing bubble crash. The data exhibits a seasonal pattern, with employment increasing in the first half of the year, decreasing afterward, and then rising again. The lag plot indicates a strong positive correlation at all lags. For a clearer seasonal decomposition, adjusting the employment numbers by the factor of population growth could be beneficial.

```{r,warning=FALSE}
aus_production %>% 
  autoplot(Bricks) +
  ggtitle("Autoplot")

aus_production %>% 
  gg_season(Bricks) +
  ggtitle("Seasonal Decomposition")

aus_production %>% 
  gg_subseries(Bricks) +
  ggtitle("Subseries Plot")

aus_production %>% 
  gg_lag(Bricks) +
  ggtitle("Lag Plot")

aus_production %>% 
  ACF(Bricks) %>% 
  autoplot() + 
  ggtitle("Autocorrelation Function")
```

For the AUS Production dataset, brick production lacks a clear trend but displays strong annual seasonality with a cyclical pattern. Production notably dropped in the early 1980s. The seasonal plot shows increases in Q1 and Q3, with a decline in Q4. The lag plot reveals consistent positive season-to-season correlation.


```{r,warning=FALSE}
pelt %>% 
  autoplot(Hare) +
  ggtitle("Autoplot")

pelt %>% 
  gg_subseries(Hare)+
  ggtitle("Subseries Plot")


pelt %>% 
  gg_subseries(Hare)+
  ggtitle("Subseries Plot")


pelt %>% 
  ACF(Hare) %>% 
  autoplot() + 
  ggtitle("Autocorrelation Function")

```

For the Pelts data set for Hare, I notice that there isn’t a distinct trend, but it is shown a potential seasonal pattern accompanied by some cyclic behavior.There seem to be sharp fluctuations in the number of traded Hare pelts through a few year periods, with a general decrease as the decade comes to an end. The lag plot illustrates a moderate positive correlation particularly in lag 1.



```{r,warning=FALSE}
PBS %>% 
  filter(ATC2 == "H02")  %>% 
  autoplot(Cost) + 
  ggtitle("Autoplot")


```

The Hare pelts dataset shows no clear trend but hints at seasonality and cyclic behavior. There are sharp fluctuations in traded pelts, with a general decline towards the end of the decade. The lag plot shows moderate positive correlation, especially at lag 1.

```{r,warning=FALSE}
us_gasoline %>% 
  autoplot() + 
  ggtitle("Autoplot")

us_gasoline %>% 
  gg_season() +
  ggtitle("Seasonal Decomposition")

us_gasoline %>% 
  gg_subseries()+
  ggtitle("Subseries Plot")

us_gasoline %>% 
  gg_lag() +
  ggtitle("Lag Plot")

us_gasoline %>% 
  ACF() %>% 
  autoplot() + 
  ggtitle("Autocorrelation Function")
```

The Gasoline Barrels series shows a positive trend with seasonal patterns but is quite noisy, with peaks and declines at specific times of the month. The lag plot reveals a positive correlation with some overplotting. No unusual years are evident, though overplotting may obscure such trends.